---
output:
  github_document:
    html_preview: false
    df_print: tibble
---

<!-- index.md is generated from index.Rmd Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/index-",
  out.width = "100%",
  message = FALSE, warning = FALSE
)
library(modifiedfisher)
```

# modifiedfisher

A non-conservative, size-α modified Fisher exact test for comparing two
proportions. Its p-value, confidence interval, and power all agree with one
another by construction, and it controls the Type I error rate strictly while
recovering most of the power that the ordinary exact test gives up to
conservativeness. It implements the test of van der Meulen, Raymond and van der
Meulen (2021), building on van der Meulen (2008).

## Installation

`modifiedfisher` is not yet on CRAN. Install the current development version
from GitHub using `devtools`:
 
```r
# install.packages("devtools")
devtools::install_github("pvdmeulen/modifiedfisher")
```
 
A CRAN submission is planned once the package clears the remaining `R CMD check`
notes. In regulated or locked-down environments where non-CRAN packages cannot
be installed, contact the maintainer for a source tarball.

## Usage

The main function takes the two success counts (`u`, `v`), the two group sizes
(`m`, `n`), and a null odds ratio, and returns a familiar `htest` object.

```{r basic, eval = FALSE}
# Example 1 of the paper: 5/12 successes versus 7/11, H0: OR = 1, alpha = 0.05.
res <- modified_fisher_exact_test(u = 5, m = 12, v = 7, n = 11, odds_ratio = 1)

res$estimate    # sample odds ratio
res$p.value     # two-sided, test-based p-value
res$conf.int    # agreeing (1 - alpha) confidence interval
```

## Learn more

- [Get started](articles/modifiedfisher.html): an introduction to `modified_fisher_exact_test()`, its arguments, power, and diagnostics.
- [Background and comparison](articles/background-and-comparison.html): why the
  test exists, and a comparison with `fisher.test()` and
  `exact2x2::fisher.exact()` (both similar to SAS PROC Freq).
- [Overview of algorithm](articles/overview-of-algorithm.html): a step by step overview of how the test is built in R.
- [Reproducing the paper's figures](articles/reproducing-paper-figures.html):
  rebuilding the size and power curves, and adapting them to your own design.

## Reference

van der Meulen EA, Raymond K, van der Meulen PJ (2021). *Consistent Confidence
Limits, P Values, and Power of the Non-Conservative, Size-α Modified Fisher
Exact Test.* Journal of Biostatistics and Biometric Applications 6(1):102.

van der Meulen EA (2008). A Nonrandomized, Nonconservative Version of the
Fisher Exact Test. *Communications in Statistics - Theory and Methods*,
37:699-708.
