---
output:
  github_document:
    df_print: tibble
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# modifiedfisher: Non-Conservative Size-α Modified Fisher Exact Test

<!-- badges: start -->
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R-CMD-check](https://github.com/pvdmeulen/modifiedfisher/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pvdmeulen/modifiedfisher/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

```{r options, echo = FALSE, message = FALSE, include = FALSE}
options(cli.ignore_unknown_rstudio_theme = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

The `modifiedfisher` package implements the non-randomised, non-conservative,
size-α modified Fisher exact test (MFET) for comparing two binomial proportions,
as introduced by van der Meulen (2008) and extended with agreeing test-based
p-values, confidence intervals, and power in van der Meulen, Raymond, and van
der Meulen (2021).

The classical Fisher exact test is the uniformly most powerful unbiased (UMPU)
test for two independent binomial proportions, but requires randomisation at the
critical values to achieve size α exactly — making it impractical without a
natural p-value or confidence interval. The standard non-randomised alternatives
are unsatisfactory: rejecting only at strictly extreme values is needlessly
conservative (as in the SAS Proc FREQ exact test), while the mid p-value and
Woolf's asymptotic test do not strictly control the Type I error rate. The MFET
resolves this by additionally rejecting when the randomisation probability at a
critical value exceeds a data-driven threshold γ₀, chosen to bring the actual
size as close to α as possible without exceeding it. P-values and confidence
intervals are both derived directly from this test, so they agree by
construction.

## Installation

```{r installation, echo = TRUE, eval = FALSE}
# Install from GitHub:
devtools::install_github("pvdmeulen/modifiedfisher")
```

## Quick start

The main function is `modified_fisher_exact_test()`. It takes the number of
successes (`u`, `v`) and number of trials (`m`, `n`) in each group, and the
odds ratio under the null hypothesis (`odds_ratio`):

```{r example1, echo = TRUE, eval = TRUE}
library(modifiedfisher)

# Example 2 from Table 2 in van der Meulen et al. (2021):
# 13/41 vs 6/47, testing H0: OR = 1
result <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                     odds_ratio = 1)
result
```

The returned object is of class `htest`, so standard components are accessible
directly:

```{r htest_components, echo = TRUE, eval = TRUE}
result$estimate   # odds ratio estimate
result$p.value    # two-sided p-value
result$conf.int   # 95% confidence interval
```

These match Table 2 of the paper (OR = 3.173, p = 0.034, 95% CI 1.082–10.25).

## Comparing with other tests

The table below reproduces Table 2 from the paper, comparing the MFET against
Woolf's asymptotic test and the SAS Proc FREQ exact test across four worked
examples. The disagreement between the SAS Proc FREQ p-value and confidence
interval in Example 2 (p < 0.05 but CI includes OR = 1) illustrates the
consistency advantage of the MFET's test-based approach.

| Example | Test | OR | p-value | 95% CI |
|---|---|---|---|---|
| 5/12 vs 7/11 | Modified FE | 0.408 | 0.321 | 0.716 – 2.210 |
| | Woolf | 0.408 | 0.296 | 0.760 – 2.193 |
| | SAS Proc FREQ | 0.408 | 0.414 | 0.550 – 2.871 |
| 13/41 vs 6/47 | Modified FE | 3.173 | **0.034** | **1.082 – 10.25** |
| | Woolf | 3.173 | 0.036 | 1.077 – 9.34 |
| | SAS Proc FREQ | 3.173 | 0.039 | 0.969 – 11.31 |
| 37/65 vs 28/71 | Modified FE | 2.029 | 0.0439 | 1.023 – 4.083 |
| | Woolf | 2.029 | 0.0425 | 1.024 – 4.021 |
| | SAS Proc FREQ | 2.029 | 0.0584 | 0.970 – 4.258 |
| 72/128 vs 58/142 | Modified FE | 1.804 | 0.0175 | 1.108 – 2.936 |
| | Woolf | 1.804 | 0.0167 | 1.113 – 2.925 |
| | SAS Proc FREQ | 1.804 | 0.0203 | 1.082 – 3.018 |

## Diagnosing size control

Setting `local_size_data = TRUE` attaches a data frame of the local size,
evaluated across the nuisance parameter space, to the returned object. This is
the diagnostic plot recommended in the paper for confirming that the size
optimisation has not been trapped at a local maximum:

```{r size_plot_data, echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE}
result_diag <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                          odds_ratio = 1,
                                          local_size_data = TRUE)

alpha <- 0.05

head(result_diag$local.size.data)
```

```{r size_plot, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE}
library(ggplot2)
library(svglite)

# A custom variant of theme_bw(), replicating ggdark::dark_theme_bw()
# without depending on the (CRAN-archived) ggdark package. This inverts
# the light/dark elements of theme_bw(): white text/lines on a dark background,
# with grid lines a shade lighter than the background for contrast.
theme_bw_dark <- function(base_size = 11, base_family = "") {
  theme_bw(base_size = base_size, base_family = base_family) %+replace%
    theme(
      panel.background = element_rect(fill = "#0D1116", colour = NA),
      plot.background  = element_rect(fill = "#0D1116", colour = NA),
      legend.background = element_rect(fill = "#0D1116", colour = NA),
      legend.key = element_rect(fill = "#0D1116", colour = NA),
      panel.border = element_rect(fill = NA, colour = "grey60"),
      panel.grid = element_line(colour = "grey25"),
      axis.text = element_text(colour = "grey85"),
      axis.title = element_text(colour = "grey95"),
      axis.ticks = element_line(colour = "grey60"),
      plot.title = element_text(colour = "grey95"),
      plot.subtitle = element_text(colour = "grey85"),
      legend.text = element_text(colour = "grey85"),
      legend.title = element_text(colour = "grey95")
    )
}

plot <- ggplot(data = result_diag$local.size.data, aes(x = pi1, y = size)) +
    geom_line(colour = "#DB5461", linewidth = 1) +
    coord_cartesian(ylim = c(0, 1.05 * alpha * 100)) +
    labs(
      title = "Size of Modified Fisher Exact Test",
      subtitle = latex2exp::TeX(paste0("\\textit{$m=", 41, ",\\,n=", 47,
                                       ",\\,H_{0}=", 1, "$}")),
      y = "Size (%)",
      x = bquote(pi[.("1")])
    ) +
    theme(
      text = element_text(size = 12),
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(family = "STIX Two Text", face = "italic"),
      axis.title.x = element_text(family = "STIX Two Text", face = "italic"),
      legend.position = "bottom",
      legend.title = element_blank(),
      legend.text = element_text(size = 12),
      plot.background = element_rect(fill = NA),
      panel.background = element_rect(fill = NA)
    )

light_plot <- plot +
  theme_bw() +
  geom_hline(yintercept = alpha * 100, linetype = 2, colour = "grey30",
             linewidth = 0.4)

dark_plot <- plot +
  theme_bw_dark() +
  geom_hline(yintercept = alpha * 100, linetype = 2, colour = "grey70",
             linewidth = 0.4)

ggsave(light_plot, filename = "man/figures/README-light_plot_data-1.svg",
       width = 160, height = 120, units = "mm", bg = "transparent",
       dpi = "retina")

ggsave(dark_plot, filename = "man/figures/README-dark_plot_data-1.svg",
       width = 160, height = 120, units = "mm", bg = "#0D1116",
       dpi = "retina")
```

<picture>
  <source media="(prefers-color-scheme: dark)" srcset="man/figures/README-dark_plot_data-1.svg">
  <source media="(prefers-color-scheme: light)" srcset="man/figures/README-light_plot_data-1.svg">
  <img alt="Plot of the local size of the modified Fisher exact test as a function of the nuisance parameter pi1, with a horizontal reference line at alpha = 0.05. The size reaches but does not exceed the reference line over a broad central region.">
</picture>

The size should reach but not exceed the dashed α reference line, confirming
non-conservative size control across the full range of the nuisance parameter.

## Learn more

The package vignette covers additional features in detail:

- Testing a null hypothesis other than OR = 1
- Controlling numerical precision and choosing the optimisation method
- Power calculations, including for superiority testing
- Accessing the test frame (critical values and randomisation probabilities)
- Reproducing the size and power comparison plots from the paper using the
  exported `local_size_*()` and `power_*()` functions

```{r vignette, echo = TRUE, eval = FALSE}
vignette("modifiedfisher")
```

## References

van der Meulen EA, Raymond K, van der Meulen PJ (2021). Consistent Confidence
Limits, P Values, and Power of the Non-Conservative, Size-α Modified Fisher
Exact Test. *Journal of Biostatistics and Biometric Applications*, 6(1):102.
<https://doi.org/10.13140/RG.2.2.34661.73448>

van der Meulen EA (2008). A Nonrandomized, Nonconservative Version of the
Fisher Exact Test. *Communications in Statistics — Theory and Methods*,
37:699–708.
