---
title: "Introduction to modifiedfisher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to modifiedfisher}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(modifiedfisher)
```

The **non-conservative, size-$\alpha$ modified Fisher exact test** (MFET)
rejects the null at a critical value only when the corresponding randomisation
probability exceeds a threshold $\gamma_0$, chosen so the test's size (maximised
over the nuisance parameter) is as large as possible without exceeding $\alpha$.
This makes it less conservative than the standard Fisher exact test while still
strictly controlling the Type I error rate. For why this matters and how it
compares to other tests, see the
[Background and comparison](background-and-comparison.html) article; for how it
is built, see [Overview of algorithm](overview-of-algorithm.html).

This vignette is a tour of `modified_fisher_exact_test()`, using the worked
examples from van der Meulen, Raymond and van der Meulen (2021).

## Basic usage

The test compares two independent binomial proportions $u/m$ and $v/n$, testing
$H_0: \theta = \theta_0$ where $\theta$ is the log odds ratio. By default
`odds_ratio = 1` (equality of proportions).

We reproduce Example 2 from the paper's Table 2: 13 of 41 versus 6 of 47.

```{r}
result <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                     odds_ratio = 1, local_size_data = TRUE)
result
```

The result is an `htest`, so the usual components are available directly, and
they match Table 2 (odds ratio 3.173, p-value 0.034, 95% CI 1.082 to 10.25):

```{r}
result$estimate
result$p.value
result$conf.int
```

### P-values and confidence intervals agree

Both are derived from the same test, so `odds_ratio` is rejected at level
`alpha` if and only if it falls outside `conf.int`. The two can never disagree.
This is not true of `stats::fisher.test()` or the SAS Proc FREQ exact test,
whose p-values and intervals are built separately and can conflict. The
[Background and comparison](background-and-comparison.html) article shows a
worked case.

## A different null hypothesis

`odds_ratio` sets $\theta_0$ and need not be 1. The estimate and confidence
interval do not depend on it; only the p-value and `$null.value` change.

```{r}
result_or2 <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                         odds_ratio = 2)
result_or2$p.value
result_or2$conf.int
```

## Precision and optimisation method

Two arguments control the numerics:

- `precision` is the tolerance for the bisection searches behind the confidence
  limits and p-value (default `1e-3`). A smaller value is more accurate but
  needs more size evaluations, which is the expensive step.
- `method` selects how the size is maximised over the nuisance parameter. The
  default `"zoom"` refines a grid search and is usually fastest; `"trust"` uses
  a trust-region optimiser with an analytic gradient. They agree to within
  `precision`. For `"zoom"`, `maze` sets the grid points per iteration and
  `zoom_iter` the number of refinements.

```{r}
result_trust <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                           odds_ratio = 1, method = "trust")
result_trust$p.value
```

## Power

By default the test also returns its power at two response rates `power_at_pi1`
and `power_at_pi2`, under the null `odds_ratio` used to build the test frame.
This is the unconditional probability of rejecting $H_0$ at those true rates and
sample sizes.

```{r}
result_power <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                           odds_ratio = 1,
                                           power_at_pi1 = 0.3, power_at_pi2 = 0.6)
result_power$power
```

Set `power = FALSE` to skip this, which saves time for larger samples. Setting
`superiority = TRUE` restricts the calculation to a positive observed effect
(one-directional alternative), which the paper notes changes power only slightly
for relevant effect sizes.

```{r}
result_sup <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                         odds_ratio = 1,
                                         power_at_pi1 = 0.3, power_at_pi2 = 0.6,
                                         superiority = TRUE)
result_sup$power
```

## Diagnostic: plotting the size

The size of the MFET is found numerically, so it is worth confirming the
optimiser has not mistaken a local peak for the global one. `local_size_data =
TRUE` attaches the local size at 101 values of the nuisance parameter $\pi_1$:

```{r}
head(result$local.size.data)
```

```{r, fig.alt = "Local size of the modified Fisher exact test against the nuisance parameter, with a reference line at alpha."}
plot(size ~ pi1, data = result$local.size.data,
     type = "l", col = "firebrick", lwd = 2,
     xlab = expression(pi[1]), ylab = "Size (%)")
abline(h = 0.05 * 100, lty = 2)
```

The curve should stay at or below $\alpha \times 100\%$ throughout, touching it
across a broad central region. That is the non-conservative behaviour behind the
MFET's power advantage.

## Internals: gamma0 and the test frame

For advanced use or for checking against the original SAS/IML macro, the result
exposes the threshold and the full test frame. `$gamma0` is the optimal
$\gamma_0$; `$support.data` holds, for every total $T$, the critical values
$c_1, c_2$ and randomisation probabilities $\gamma_1, \gamma_2$.

```{r}
result$gamma0
head(result$support.data)
```

The boundary outcomes the MFET rejects (but the conservative test does not) are
exactly those with a randomisation probability above the threshold:

```{r}
subset(result$support.data, gamma1 > result$gamma0 | gamma2 > result$gamma0)
```

The mechanics of this are covered in
[Overview of algorithm](overview-of-algorithm.html), and the size and power of
the comparison tests in
[Reproducing the paper's figures](reproducing-paper-figures.html).

## References

van der Meulen EA, Raymond K, van der Meulen PJ (2021). Consistent Confidence
Limits, P Values, and Power of the Non-Conservative, Size-$\alpha$ Modified
Fisher Exact Test. *Journal of Biostatistics and Biometric Applications*,
6(1):102.

van der Meulen EA (2008). A Nonrandomized, Nonconservative Version of the Fisher
Exact Test. *Communications in Statistics - Theory and Methods*, 37:699-708.
