---
title: "Get started"
subtitle: "An overview of modifiedfisher and its options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(dplyr)
library(ggplot2)
```

The **non-conservative, size-$\alpha$ modified Fisher exact test** rejects the null at a critical value only when the corresponding randomisation probability exceeds a threshold $\gamma_0$, chosen so the test's size (maximised over the nuisance parameter) is as large as possible without exceeding $\alpha$.

This makes it less conservative than the standard Fisher exact test while still
strictly controlling the Type I error rate. For why this matters and how it
compares to other tests, see the [Background and
comparison](background-and-comparison.html) article; for how it is built, see
[Overview of algorithm](overview-of-algorithm.html).

This vignette is a tour of `modified_fisher_exact_test()`, using the worked
examples from van der Meulen, Raymond and van der Meulen (2021).

## Basic usage

The test compares two independent binomial proportions $u/m$ and $v/n$, testing
$H_0: \theta = \theta_0$ where $\theta$ is the log odds ratio. By default
`odds_ratio = 1` (equality of proportions).

We reproduce Example 2 from the paper's Table 2: 13 of 41 versus 6 of 47.

```{r}
library(modifiedfisher)
result <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                     odds_ratio = 1, local_size_data = TRUE)
result
```

The result is an `htest`, so the usual components are available directly, and
they match Table 2 (odds ratio 3.173, p-value 0.034, 95% CI 1.082 to 10.25):

```{r}
result$estimate
result$p.value
result$conf.int
```

### P-values and confidence intervals agree

Both are derived from the same test, so `odds_ratio` is rejected at level
`alpha` if and only if it falls outside `conf.int`. The two can never disagree.
This is not true of `stats::fisher.test()` or the SAS Proc FREQ exact test,
whose p-values and intervals are built separately and can conflict. The
[Background and comparison](background-and-comparison.html) article shows a
worked case.

### Scope and limitations
 
`modified_fisher_exact_test()` implements a two-sided test only. The null
hypothesis is $H_0$: OR = `odds_ratio`, and the alternative is two-sided:
OR $\neq$ `odds_ratio`. One-sided p-values and one-sided confidence bounds are
not currently returned.
 
For superiority trials where the direction of the effect is specified in
advance, the `superiority = TRUE` argument adjusts the **power calculation**
to count only tables where the observed rate in group 2 exceeds that in group 1.
This does not change the test or p-value; it narrows the power evaluation to
the relevant region.
 
Formal one-sided inference (a one-sided p-value and a one-sided confidence
bound) is outside the current scope of the package.

## A different null hypothesis

The `odds_ratio` argument sets $\theta_0$; this does not need be 1. The estimate and confidence interval do not depend on it (only the p-value and `$null.value` change).

```{r}
result_or2 <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                         odds_ratio = 2)
result_or2$p.value
result_or2$conf.int
```

## Precision and optimisation method

Two arguments control the numerics:

- `precision` is the tolerance for the bisection searches behind the confidence limits and p-value (default `1e-3`). A smaller value is more accurate but needs more size evaluations, which is the expensive step.
  
- `method` selects how the size is maximised over the nuisance parameter. The default `"zoom"` refines a grid search and is usually fastest; `"trust"` uses a trust-region optimiser with an analytic gradient. They agree to within `precision`. For `"zoom"`, `maze` sets the grid points per iteration and `zoom_iter` the number of refinements.

```{r}
result_trust <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                           odds_ratio = 1, method = "trust")

result_trust$p.value
```

## Power

The test also returns its power at two response rates `power_at_pi1`
and `power_at_pi2` under the null `odds_ratio` used to build the test frame. This is the unconditional probability of rejecting $H_0$ at when $\pi_1=$`power_at_pi1` and $\pi_2=$`power_at_pi2`. The default values are 0.5 and 0.75, respectively, but these can be changed:

```{r}
result_power <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                           odds_ratio = 1, power_at_pi1 = 0.3,
                                           power_at_pi2 = 0.6)

result_power$power
```

Set `power = FALSE` to skip this, which saves time for larger samples. Setting
`superiority = TRUE` restricts the calculation to a positive observed effect
(one-directional alternative), which (as the paper notes) only slightly changes power for relevant effect sizes:

```{r}
result_sup <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                         odds_ratio = 1, power_at_pi1 = 0.3,
                                         power_at_pi2 = 0.6, superiority = TRUE)

result_sup$power
```

## Sample size planning
 
The power functions can be inverted to answer the planning question: given a
target power, what sample sizes are needed? There is no closed-form answer, so
the simplest approach is a grid search over candidate values of $m$ and $n$.
 
The example below finds the smallest equal-allocation sample size ($m = n$)
needed to achieve 80% power against $\pi_1 = 0.3$, $\pi_2 = 0.6$ under
$H_0$: OR = 1 at $\alpha = 0.05$:
 
```{r sample-size, cache = TRUE}
target_power <- 80   # percent
pi1 <- 0.3
pi2 <- 0.6
alpha <- 0.05
 
# Search over candidate n = m from 10 to 100
results <- data.frame(n = integer(), power = numeric())
 
for (n_candidate in 10:100) {
  
  df_candidate <- construct_test_frame(
    .odds_ratio = 1, .m = n_candidate, .n = n_candidate,
    .alpha = alpha, .precision = 1e-3
  )
  
  g0_candidate <- optimise_gamma0(
    .odds_ratio = 1, .m = n_candidate, .n = n_candidate,
    .alpha = alpha, .precision = 1e-3,
    .method = "zoom", .maze = 10, .zoom_iter = 6
  )
  
  pwr <- power_modified(
    p          = c(pi1, pi2),
    .gamma0    = g0_candidate,
    .odds_ratio = 1,
    .m         = n_candidate,
    .n         = n_candidate,
    .df        = df_candidate,
    .alpha     = alpha,
    .precision = 1e-3,
    .superiority = FALSE
  )
  
  results <- rbind(results, data.frame(n = n_candidate, power = pwr * 100))
  
  # Stop when desired power is reached:
  if (pwr * 100 >= target_power) break
  
}
 
# Show last three rows:
tail(results, 3)
```
 
The smallest $n$ achieving 80% power is `r results$n[nrow(results)]` per
group, giving `r round(results$power[nrow(results)], 1)`% power. The search
stops at the first $n$ that crosses the threshold; inspect `results` to see
how power grows with sample size.
 
For unequal allocation, replace the loop with a grid over pairs $(m, n)$.
Note that the power calculation re-runs the full test optimisation at each
candidate sample size, so the search is slow for large candidate ranges. A
coarse first pass (step size 5 or 10) followed by a fine search around the
crossing point is more efficient.

## Diagnostic: plotting the size

The size of the modified test is found numerically, so it is worth confirming
the optimiser has not mistaken a local peak for the global one. `local_size_data
= TRUE` attaches the local size at 101 values of the nuisance parameter $\pi_1$:

```{r}
result$local.size.data |> head()
```

Plotting this data:

```{r, fig.alt = "Local size of the modified Fisher exact test against the nuisance parameter, with a reference line at alpha.", echo = FALSE}
plotdata <- result$local.size.data

p <- ggplot(
  data = plotdata,
  aes(x = pi1, y = size)
  ) +
  geom_line(colour = "#016c59", linewidth = 0.8, linetype = 1) +
  geom_hline(yintercept = 5, linetype = 2, colour = "grey40") +
  labs(x = expression(pi[1]), y = "Size (%)") +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    plot.background = element_rect(fill = "transparent")
  )

print(p)

#plot(size ~ pi1, data = result$local.size.data,
#     type = "l", col = "firebrick", lwd = 2,
#     xlab = expression(pi[1]), ylab = "Size (%)")
#abline(h = 0.05 * 100, lty = 2)
```

The curve should stay at or below $\alpha \times 100\%$ throughout, touching it
across a broad central region. This is the non-conservative behaviour behind the
modified test's power advantage.

## Internals: gamma0 and the test frame

For advanced use or for checking against the original SAS/IML macro, the result
exposes the threshold and the full test frame. `$gamma0` is the optimal
$\gamma_0$; `$support.data` holds, for every total $T$, the critical values
$c_1, c_2$ and randomisation probabilities $\gamma_1, \gamma_2$.

```{r}
result$gamma0
result$support.data |> head()
```

The boundary outcomes the modified test rejects (but the conservative test does
not) are exactly those with a randomisation probability above the threshold:

```{r}
result$support.data |> 
  filter(gamma1 > result$gamma0 | gamma2 > result$gamma0)
```

The mechanics of this are covered in
[Overview of algorithm](overview-of-algorithm.html), and the size and power of
the comparison tests in
[Reproducing the paper's figures](reproducing-paper-figures.html).

## References

van der Meulen EA, Raymond K, van der Meulen PJ (2021). Consistent Confidence
Limits, P Values, and Power of the Non-Conservative, Size-$\alpha$ Modified
Fisher Exact Test. *Journal of Biostatistics and Biometric Applications*,
6(1):102.

van der Meulen EA (2008). A Nonrandomized, Nonconservative Version of the Fisher
Exact Test. *Communications in Statistics - Theory and Methods*, 37:699-708.
