---
title: "Introduction to modifiedfisher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to modifiedfisher}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(modifiedfisher)
```

## Introduction

The classical Fisher exact test is the uniformly most powerful unbiased
(UMPU) test for comparing two binomial proportions, but its UMPU property
relies on randomisation at the critical values. A non-randomised test is
of more practical use, but the usual non-randomised versions either fail to
control the Type I error rate strictly (e.g. the mid p-value, or Woolf's
asymptotic test) or are needlessly conservative (e.g. the SAS Proc FREQ
exact test).

The **non-conservative, size-$\alpha$ modified Fisher exact test**
(MFET) addresses this gap. It is derived from the randomised Fisher exact
test by rejecting the null at a critical value only if the corresponding
randomisation probability exceeds a threshold $\gamma_0$. The value of
$\gamma_0$ is chosen so that the actual size of the test, maximised over
the nuisance parameter, is as large as possible while remaining no greater
than $\alpha$. This makes the test less conservative than the standard
Fisher exact test while still strictly controlling the Type I error rate.

This vignette walks through the main features of `modified_fisher_exact_test()`
using the worked examples from van der Meulen, Raymond and van der Meulen
(2021), *Consistent Confidence Limits, P Values, and Power of the
Non-Conservative, Size-$\alpha$ Modified Fisher Exact Test*.

## Basic usage

The test compares two independent binomial proportions, $u/m$ and
$v/n$. The null hypothesis is $H_0: \theta = \theta_0$, where
$\theta$ is the log odds ratio. By default, `odds_ratio = 1`, i.e. the
test is for equality of proportions.

We reproduce Example 2 from Table 2 in the original paper: 13 successes out of
41 trials in group 1, versus 6 successes out of 47 trials in group 2.

```{r}
result <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47, odds_ratio = 1, local_size_data = TRUE)

result
```

The returned object is of class `htest`, so the familiar components
(`$p.value`, `$estimate`, `$conf.int`) are all available directly:

```{r}
result$estimate

result$p.value

result$conf.int
```

These match Table 2 of the paper: an odds ratio estimate of 3.173, a two-sided p-value of 0.034, and a 95% confidence interval of (1.082, 10.25).

### Agreement between p-values and confidence intervals

A central feature of the MFET is that its p-values and confidence intervals are constructed to *agree by definition*, because both are derived from the same underlying test. The null hypothesis value `odds_ratio` is rejected at level `alpha` if and only if it falls outside the `conf.int`:

```{r}
alpha <- 0.05

rejects_null <- result$p.value < alpha
one_outside_ci <- !(1 >= result$conf.int[1] && 1 <= result$conf.int[2])

rejects_null
one_outside_ci
```

This is in contrast to, for example, the SAS Proc FREQ exact test, where the p-value is based on a sum of conditional tail probabilities but the confidence interval is constructed separately using equal-tail one-sided p-values, meaning these can disagree (see Table 2 Example 2 in the paper, where the SAS Proc FREQ confidence interval includes an odds ratio of 1 despite a p-value below 0.05).

## Testing a different null hypothesis

The `odds_ratio` argument specifies $\theta_0$, the value of the odds ratio under $H_0$. It need not be 1. For example, to test whether the odds ratio differs from 2:

```{r}
result_or2 <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47, odds_ratio = 2)

result_or2$p.value
result_or2$conf.int
```

The point estimate and confidence interval do not depend on `odds_ratio`: only the p-value (which tests $H_0: \theta = \theta_0$) and the `$null.value` element change.

## Controlling precision and choosing the optimisation method

Two arguments control the numerical behaviour of the test:

- `precision` sets the tolerance for the bisection searches used to find
  confidence limits and p-values (default `1e-03`).
- `method` selects how the actual size of the test is maximised over the nuisance parameter at each step of those bisection searches. The default, `"zoom"`, iteratively refines a grid search and is generally fastest. The alternative, `"trust"`, uses a trust-region optimiser (from the **trust** package) with an analytic gradient.

```{r}
result_trust <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47, odds_ratio = 1, method = "trust")

result_trust$p.value
result_trust$conf.int
```

The two methods should agree to within `precision`. Increasing the `precision` (i.e. setting a lower value like `1e-6`) increases the accuracy of the bisection at the cost of additional evaluations of the test's size, which is the most computationally expensive step. The `maze` and `zoom_iter` arguments further control the `"zoom"`vmethod specifically: `maze` is the number of grid points evaluated at each iteration, and `zoom_iter` is the number of refinement iterations.

## Power calculations

By default, `modified_fisher_exact_test()` also returns the power of the test, evaluated at two user-specified response rates `power_at_pi1` and `power_at_pi2` (defaults to 0.5 and 0.75, respectively), under the null `odds_ratio` used to build the test frame (typically `odds_ratio = 1`):

```{r}
result_power <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47, odds_ratio = 1, power_at_pi1 = 0.3, power_at_pi2 = 0.6)

result_power$power
```

This is the unconditional probability of rejecting $H_0:$ OR = 1 when the true response rates are 0.3 in group 1 and 0.6 in group 2, for the sample sizes `m = 41` and `n = 47` used in this call.

If the power calculation is not needed, setting `power = FALSE` skips it, which can meaningfully reduce computation time for larger sample sizes.

### Testing superiority

Setting `superiority = TRUE` restricts the power calculation to the sub-region where the observed response rate in group 2 exceeds that in group 1 (i.e. a positive observed effect), which is appropriate when the alternative hypothesis of interest is one-directional:

```{r}
result_superiority <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47, odds_ratio = 1, power_at_pi1 = 0.3, power_at_pi2 = 0.6, superiority = TRUE)

result_superiority$power
```

As noted in the paper, this restriction typically has only a small effect on power for clinically relevant effect sizes.

## Diagnosing the test: plotting the size

Because the size of the MFET is determined numerically (by maximising the local size over the nuisance parameter, subject to it not exceeding `alpha`), the paper recommends plotting the size as a function of the nuisance parameter as a diagnostic check (particularly to confirm that no local optimum has been mistaken for a global optimum).

Setting `local_size_data = TRUE` (like in the earlier function call) attaches a data frame of the local size, evaluated at 101 equally spaced values of the nuisance parameter $p_0 \in [0, 1]$, to the returned object:

```{r}
head(result$local.size.data)
```

This can be plotted directly:

```{r, fig.alt = "Plot of the local size of the modified Fisher exact test as a function of the nuisance parameter pi1, with a horizontal reference line at alpha = 0.05"}
plot(size ~ pi1, data = result$local.size.data,
     type = "l", col = "firebrick", lwd = 2,
     xlab = expression(pi[1]), ylab = "Size (%)",
     main = "Size of the modified Fisher exact test")
abline(h = alpha * 100, lty = 2)
```

The size should remain at or below the dashed reference line ($\alpha \times 100\%$) across the full range of $\pi_1$, reaching it (but not exceeding it) over a broad central region; this is the non-conservative behaviour that gives the MFET its power advantage over, for example, the Proc FREQ exact test in SAS.

## The gamma0 threshold and the test frame

Two further elements of the returned object give access to the internal machinery of the test, which may be of interest for advanced use or for verifying results against the original SAS/IML macro.

`$gamma0` is the optimal threshold $\gamma_0$: a critical value is only rejected if its randomisation probability exceeds this threshold.

```{r}
result$gamma0
```

`$support.data` is the full test frame: for every possible total $T = 0, \ldots, m+n$, the critical values $c_1, c_2$ and the randomisation probabilities $\gamma_1, \gamma_2$ that define the randomised Fisher exact test from which the MFET is derived.

```{r}
head(result$support.data)
```

A critical value is included in the MFET's rejection region if and only if its randomisation probability exceeds `$gamma0`. For example, the boundary values rejected by the MFET (but not by the conservative non-randomised Fisher exact test) are those where `gamma1 > gamma0` or `gamma2 > gamma0`:

```{r}
gamma0 <- result$gamma0
df <- result$support.data

subset(df, gamma1 > gamma0 | gamma2 > gamma0)
```

## Comparison with other tests

The paper compares the MFET against three alternatives: the conservative (non-randomised) Fisher exact test, Woolf's asymptotic test, and the SAS Proc FREQ exact test (based on summing conditional tail probabilities).

Helper functions for the size and power of these alternatives are also available in this package (accessed through `modifiedfisher:::`):

- `local_size_woolf()` and `power_woolf()` for Woolf's test, and
- `local_size_procfreq()` and `power_procfreq()` for the SAS Proc FREQ exact test.

These are primarily intended for reproducing the size and power comparison plots in the paper (Figures 1-6), and are documented separately.

## Summary

`modified_fisher_exact_test()` provides:

- A two-sided p-value for $H_0: \theta = \theta_0$, for any $\theta_0$ (not just $\theta_0 = 1$).
- A two-sided $(1-\alpha)$ confidence interval for the odds ratio, guaranteed to agree with the p-value.
- The power of the test at user-specified response rates, optionally restricted to a superiority sub-region.
- Diagnostic data (`local.size.data`) for confirming size control across the nuisance parameter space.

All of this is achieved while strictly controlling the Type I error rate at the nominal level `alpha`, in a less conservative way than the commonly used SAS Proc FREQ exact test.

## References

van der Meulen EA, Raymond K, van der Meulen PJ (2021). Consistent Confidence Limits, P Values, and Power of the Non-Conservative, Size-$\alpha$ Modified Fisher Exact Test. *Journal of Biostatistics and Biometric Applications*, 6(1):102.

van der Meulen EA (2008). A Nonrandomized, Nonconservative Version of the Fisher Exact Test. *Communications in Statistics - Theory and Methods*, 37:699-708.
