---
title: "Background and comparison"
subtitle: "The rationale behind the modified test and a comparison with other exact tests"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/background-",
  out.width = "100%",
  message = FALSE, warning = FALSE
)

# Packages:

library(modifiedfisher)
library(dplyr)
library(exact2x2)
# exact2x2 is a Suggests dependency, used only for the comparison below.
```

This article is the statistical background for the package, and a comparison
against base R's `fisher.test()` and `exact2x2::fisher.exact()`. The full
derivations are in van der Meulen, Raymond and van der Meulen (2021), building
on van der Meulen (2008).

## Why a modified test

Fisher's Exact Test fixes the total number of successes and asks only how they
split between the groups; under the null that split is non-central
hypergeometric with the odds ratio as its parameter. It never exceeds its
nominal level α, but because the data are discrete it cannot in general hit size
α exactly, so it ends up **conservative**: its true rejection rate sits below α,
and a conservative test is underpowered.

Take a worked example with two independent binomial proportions $u/m$ and $v/n$,
testing $H_0: \theta = 1$ where $\theta$ is the log odds ratio. Given $u=13$, $m=41$, $v=6$, and $n=47$ (Example 2 from the paper), the probability distribution for $u$ is given by:

```{r prob_plot, echo = FALSE, include = TRUE, eval = TRUE}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)

m_obs  <- 41                 # observed trials in Group A
n_obs  <- 47                 # observed trials in Group B
or <- 1                  # odds ratio to test

u_obs   <- 13            # observed successes in Group A
v_obs   <- 6             # observed successes in Group B
t_obs  <- u_obs + v_obs  # t = 19; the fixed total T

# Support: all valid u given T, with v = T - u:

plot_s <- seq(max(0L, t_obs - n_obs), min(m_obs, t_obs))  # 0:19
plot_t <- t_obs - plot_s                                      # 19:0

plotdata <- tibble(
  s  = plot_s,
  t  = plot_t,
  m  = m_obs,
  n  = n_obs,
  or = or
  )

plotdata <- plotdata |>
  mutate(
    prob = pmap_dbl(
      .l = list(s = s, t = t, m = m, n = n),
      .f = modifiedfisher:::pvalue_probability
      ),
    obs = case_when(
      s == u_obs ~ TRUE,
      TRUE ~ FALSE
    )
  )

p <- ggplot(
  data = plotdata,
  aes(x = s, y = prob)
  ) +
  geom_col(aes(fill = obs)) +
  scale_fill_manual(
    values = c(
      `TRUE` = "#016c59",
      `FALSE` = "#016c59"
      )
    ) +
  scale_y_continuous(
    breaks = c(0, 0.05, 0.25, 0.5, 0.75, 1)
  ) +
  geom_hline(yintercept = 0.05, linetype = 2, colour = "grey40") +
  labs(x = "u", y = "probability") +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    plot.background = element_rect(fill = "transparent")
  )

print(p)
```

Fisher's Exact Test is only *uniformly most powerful unbiased* (UMPU) if it is
allowed to **randomise** at the borderline outcomes (e.g. randomising between $u=\{3,4\}$ and $u=\{8,9\}$), in effect flipping a weighted coin to decide significance. That is rarely acceptable in practice, so a non-randomised version is used instead. The non-randomised tests all try to recover the power lost to dropping the coin flip, and differ in how, and in whether they still strictly control the Type I error rate:

- The **conservative / probability-based test** rejects on the summed probability of outcomes at least as likely as the one observed. This is the "exact" test in SAS Proc FREQ and base R's `fisher.test()`. It controls the size but is conservative, and its p-value and confidence interval can disagree (shown below).
- **Asymptotic tests** (Woolf's) and **mid-*p*** approaches are less conservative but do not strictly control the size; the true size can be larger than α.

The **modified Fisher exact test** in this package keeps strict size control
while staying closest to the randomised UMPU test. It rejects at a borderline
outcome only when its randomisation probability exceeds a single threshold γ₀,
chosen to be as large as possible while keeping the maximum size (over the
nuisance parameter) at or below α. Its p-value and confidence interval are
**test-based**, so they agree by construction. The [Overview of
algorithm](overview-of-algorithm.html) article shows how this is implemented.

## The probability-based tests in R and SAS

The probability-based tests are familiar to many practitioners.

The **two-sided p-value** uses a likelihood ordering: it sums the probabilities
of all tables no more likely than the observed one. Base R's `fisher.test()` and
SAS Proc FREQ use this same rule (in `exact2x2` it is `tsmethod = "minlike"`),
and it is the test implemented here as `pvalue_probability()`.

Its **confidence interval** is calculated using a different method.
`fisher.test()` does not invert that p-value; it uses an equal-tail (central)
construction instead, the same mismatch holds for Proc FREQ. Because the two
come from different rules, they can **disagree**: a result can be significant by
its p-value while its interval still contains an odds ratio of 1.

Two further points concern the odds-ratio **estimate**, not the test.
`fisher.test()` reports the conditional maximum likelihood estimate, whereas this
package (following the paper) reports the sample odds ratio, so the estimates
differ on identical data. And `exact2x2::fisher.exact()` lets you pick the
two-sided method *and* pairs it with the matching interval: with `tsmethod =
"minlike"` the interval inverts the same p-value, so the two agree - making it a suitable conservative comparator to `modifiedfisher`.

## Side-by-side comparison

We use the paper's Example 2: 13 of 41 in Group A versus 6 of 47 in Group B,
testing the null odds ratio of 1 at α = 0.05.

```{r calls}
tab <- matrix(c(13, 28,
                 6, 41), nrow = 2, byrow = TRUE,
              dimnames = list(Group   = c("A", "B"),
                              Outcome = c("Success", "Failure")))

mfe <- modified_fisher_exact_test(u = 13, m = 41, v = 6, n = 47,
                                  odds_ratio = 1, power = FALSE)
ft <- fisher.test(tab)
```

```{r table}
results <- tibble(
  Test       = c("Modified FE (this package)", "fisher.test() (base R)"),
  `OR est.`  = c(unname(mfe$estimate), unname(ft$estimate)),
  `p-value`  = c(mfe$p.value, ft$p.value),
  `CI lower` = c(mfe$conf.int[1], ft$conf.int[1]),
  `CI upper` = c(mfe$conf.int[2], ft$conf.int[2])
)

# exact2x2 rows are added only if that (Suggests) package is installed.
if (requireNamespace("exact2x2", quietly = TRUE)) {
  fe_min <- exact2x2::fisher.exact(tab, tsmethod = "minlike")
  fe_cen <- exact2x2::fisher.exact(tab, tsmethod = "central")
  results <- bind_rows(
    results,
    tibble(
      Test       = c("fisher.exact() minlike (exact2x2)",
                     "fisher.exact() central (exact2x2)"),
      `OR est.`  = c(unname(fe_min$estimate), unname(fe_cen$estimate)),
      `p-value`  = c(fe_min$p.value, fe_cen$p.value),
      `CI lower` = c(fe_min$conf.int[1], fe_cen$conf.int[1]),
      `CI upper` = c(fe_min$conf.int[2], fe_cen$conf.int[2])
    )
  )
}

# Does the 95% interval still contain the null odds ratio of 1?
results$`OR = 1 in CI?` <- ifelse(
  results$`CI lower` <= 1 & results$`CI upper` >= 1, "yes", "no"
)

knitr::kable(results, digits = c(NA, 3, 3, 3, 3, NA))
```

Two things stand out:

- The odds-ratio estimates differ, because `fisher.test()` and `fisher.exact()` report the conditional MLE while the modified test reports the sample odds ratio.
- The last two columns, read together, show the disagreement between p-values and confidence intervals: base R's `fisher.test()` can return a p-value below 0.05 while its 95% interval still contains an odds ratio of 1, because p-value and interval use
different rules. The `central` variant makes the equal-tail interval (and its
larger matching p-value) explicit; the `minlike` variant pairs the
likelihood-ordering p-value with an agreeing interval. The modified test, like
the agreeing `exact2x2` options, never points its p-value and interval in
opposite directions.

## Which test controls what

Strict size control here means the actual size never exceeds α for any nuisance
parameter value.

```{r properties}
properties <- tibble(
  Test = c("Randomised FE (UMPU)",
           "Conservative / probability-based (base fisher.test, Proc FREQ)",
           "exact2x2 fisher.exact, minlike",
           "Woolf asymptotic",
           "Modified FE (this package)"),
  `Strict size control?` = c("yes (size = α exactly)",
                             "yes (but conservative)",
                             "yes (but conservative)",
                             "no (can exceed α)",
                             "yes (close to α)"),
  `Non-randomised?` = c("no", "yes", "yes", "yes", "yes"),
  `p-value and CI agree?` = c("not usable", "no", "yes", "yes", "yes"),
  `OR estimate` = c("conditional MLE", "conditional MLE", "conditional MLE",
                   "sample OR", "sample OR")
)

knitr::kable(properties)
```

The randomised test is optimal but unusable in practice. The probability-based
test is reproducible and controls the size, but is conservative and can give a
disagreeing p-value and interval. Woolf's test is self-consistent and cheap but
does not strictly control the size. The modified test is the only row that is
non-randomised, strictly size-controlling, and agreeing at once, which is what
suits it to small-sample work needing strict Type I error control. The size and
power of these tests are compared across the nuisance parameter in
[Reproducing the paper's figures](reproducing-paper-figures.html).

## References

van der Meulen EA, Raymond K, van der Meulen PJ (2021). *Consistent Confidence
Limits, P Values, and Power of the Non-Conservative, Size-α Modified Fisher
Exact Test.* Journal of Biostatistics and Biometric Applications 6(1):102.

van der Meulen EA (2008). A Nonrandomized, Nonconservative Version of the
Fisher Exact Test. *Communications in Statistics - Theory and Methods*,
37:699-708.

Fay MP (2010). Confidence intervals that match Fisher's exact or Blaker's exact
tests. *Biostatistics* 11(2):373-374. (The `exact2x2` package.)
