---
title: "Overview of algorithm"
subtitle: "A step by step overview of the algorithm and the helper functions it relies on"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/algorithm-",
  out.width = "100%",
  message = FALSE, warning = FALSE
)

library(modifiedfisher)
```

This article traces how the modified Fisher exact test is built in R, from the
test frame through to the agreeing p-value and confidence interval. It assumes
the statistical background, which is in the
[Background and comparison](background-and-comparison.html) article; the
derivations and proofs are in van der Meulen, Raymond and van der Meulen (2021).

The notation: $u$ successes from $m$ trials in Group A, $v$ from $n$ in Group B,
with total $T = u + v$.

```{r contingency_table, echo = FALSE}
contingency <- tibble::tibble(
  " "                = c("Group B: Success", "Group B: Failure", "Total"),
  "Group A: Success" = c("..", "..", "$u$"),
  "Group A: Failure" = c("..", "..", "$u'$"),
  "Total"            = c("$v$",  "$v'$", "$t$")
)
knitr::kable(contingency, align = c("l", "c", "c", "c"))
```

## The basic premise

The modified test makes a single change to the randomised (UMPU) Fisher test:
instead of always rejecting at a critical value (randomised) or never rejecting
there (conservative), it uses one global threshold $\gamma_0$ and rejects at a
boundary only if that boundary's randomisation probability $\gamma$ exceeds
$\gamma_0$. The randomisation is taken away, and most of the boundary rejection
mass is recovered.

The implementation chooses $\gamma_0$ for a given null odds ratio, then inverts
the resulting test for the p-value and confidence interval. The entry point is
`modified_fisher_exact_test()`:

```{r mfet_example, eval = FALSE}
modified_fisher_exact_test(u = 5, m = 12, v = 7, n = 11, odds_ratio = 1)
```

The steps below show what the above function does internally. Helper functions
not exported to the namespace are prefixed by `.`, these can still be accessed
through `modifiedfisher:::` (e.g. for troubleshooting purposes, or to visualise
the steps with your own example as in this article).

## Step 1: Build the test frame of critical values and randomisations

For a given null odds ratio $\theta_0$ and each total $T = 0, \ldots, m+n$, find
the critical values $c_1, c_2$ and admissible randomisation probabilities
$\gamma_1, \gamma_2$. This is `construct_test_frame()`, returning one row per $T$
with columns `c1`, `c2`, `gamma1`, `gamma2` (and the starting quantiles `d1`,
`d2`).

With the critical values fixed, the two requirements (conditional size $\alpha$
and unbiasedness) form a $2\times2$ linear system solved for $(\gamma_1,
\gamma_2)$ by `.find_gamma12()`. `construct_test_frame()` searches for the
critical values giving an admissible solution (both $\gamma$'s in $[0, 1]$),
starting at the $\alpha/2$ quantiles of the non-central hypergeometric
distribution and spiralling outward in larger squares until one is found for
every $T$. At $T = 0$ and $T = m+n$ there is no room to randomise, so $\gamma_1 =
\gamma_2 = \alpha/2$. The frame holds $2(m+n+1)$ randomisation probabilities in
total, which matters in Step 4.

```{r build_frame, eval = FALSE}
# Example 1 from the paper (m = 12, n = 11):
construct_test_frame(.odds_ratio = 1, .m = 12, .n = 11, .alpha = 0.05, .precision = 1e-3)
```

The first ten rows of this dataframe are given by:

```{r show_frame, eval = TRUE, echo = FALSE}
testframe <- construct_test_frame(.odds_ratio = 1, .m = 12, .n = 11, .alpha = 0.05, .precision = 1e-3)

head(testframe, 10)
```

## Step 2: Define the rejection rule for a candidate $\gamma_0$
 
Given the frame and a candidate $\gamma_0$, the verdict on any table $(u, T)$ is
deterministic:
 
- $u$ outside $[c_1, c_2]$: reject;
- $u$ strictly inside $(c_1, c_2)$: accept;
- $u = c_1$: reject iff $\gamma_1 > \gamma_0$;
- $u = c_2$: reject iff $\gamma_2 > \gamma_0$;
- $c_1 = c_2$ (a single boundary point): reject iff $\gamma_1 + \gamma_2 > \gamma_0$.
 
This is `.modified_reject()`. Since the verdict depends only on the frame and
$\gamma_0$, not on the unknown success probability, the whole rejection region
is built once into an $(m+1)\times(n+1)$ 0/1 matrix
(`.build_rejection_matrix()`), so later steps avoid recomputing it.
 
## Step 3: Measure the (worst case) size over the nuisance parameter $\pi_1$
 
A $\gamma_0$ is acceptable only if the test's true (unconditional) size stays
below $\alpha$. Conditional on $T$ the test is size-$\alpha$ by construction,
but **unconditionally** the rejection probability depends on the unknown success
rate $\pi_1$. The real size is therefore a function of $\pi_1$, and we need to
guard against the 'worst case'.
 
For a fixed $\pi_1$, the rejection probability is a sum over all tables weighted by
their two binomial probabilities, computed as the bilinear form $p_u^\top R\,
p_v$ with $R$ the rejection matrix from Step 2 (`local_size_modified()`). The
size is the maximum over $\pi_1 \in (0, 1)$ (`size_modified()`). 

Maximisation uses either `"zoom"` (grid, then re-grid finer around the peak,
repeat; the default, and also the SAS macro's approach) or `"trust"` (a
trust-region optimiser with the analytic gradient
`.local_size_gradient_modified()`). The size curve can be multi-peaked, so the
safest check is to plot it via `local_size_data = TRUE`.
 
```{r local_size, eval = FALSE}
testframe <- construct_test_frame(.odds_ratio = 1, .m = 12, .n = 11, .alpha = 0.05, .precision = 1e-3)

# Worst-case size at a candidate gamma0 of 0.05:
size_modified(.c = 0.05, .odds_ratio = 1, .m = 12, .n = 11, .df = testframe,
              .alpha = 0.05, .precision = 1e-3, .method = "zoom",
              .maze = 10, .zoom_iter = 6)
```
 
## Step 4: Find the optimal $\gamma_0$ by bisection
 
We still want the largest worst-case size to be less than $\alpha$. The size
only changes when $\gamma_0$ crosses one of the $2(m+n+1)$ actual $\gamma$
values from Step 1, so rather than search a continuum, we sort those values and
search among them.

Raising $\gamma_0$ can only *remove* boundaries, so size is monotone in the
sorted index, which makes a bisection appropriate. This is done in
`optimise_gamma0()`: it sorts the pooled `gamma1`/`gamma2` vector and bisects to
the largest threshold whose size does not exceed $\alpha$. The optimum is not a
single number but a half-open interval between two adjacent sorted $\gamma$'s
and any value in it gives the same test.
 
```{r opt_gamma0, eval = FALSE}
optimise_gamma0(.odds_ratio = 1, .m = 12, .n = 11, .alpha = 0.05,
                .precision = 1e-3, .method = "zoom", .maze = 10, .zoom_iter = 6)
```
 
## Step 5: Invert the test for the p-value and confidence interval
 
Steps 1 to 4 decide, at some given null odds ratio $\theta_0$, whether the
observed table is rejected (and the `.accept()` helper function ties it
together: build the frame, optimise $\gamma_0$, apply the rule). Confidence
intervals and p-values are obtained by inverting that decision:
 
- **Confidence interval**: all $\theta_0$ not rejected at level $\alpha$. Its
  edges are found by bisection, starting from Woolf's asymptotic limits and
  moving inward to the first rejection, to the requested `precision`.
- **P-value**: the smallest $\alpha$ at which the table is rejected for the given
  null, found by bisecting on $\alpha$.
 
Both invert the same test, so they agree by construction. 

The main `modified_fisher_exact_test()` function runs the following pipeline:

`construct_test_frame()` → `.modified_reject()` / `.build_rejection_matrix()` → `local_size_modified()` / `size_modified()` → `optimise_gamma0()` → `.accept()`

The function returns an `htest` with `$p.value`,`$estimate`, `$conf.int`, the optimal `$gamma0`, the test frame, and optionally the size-versus-$\pi_1$ dataframe:
 
```{r mfet_full, eval = FALSE}
result <- modified_fisher_exact_test(u = 5, m = 12, v = 7, n = 11, odds_ratio = 1, local_size_data = TRUE)

result$estimate
result$p.value
result$conf.int
result$local.size.data
```

## Numerical limits and edge cases
 
**Degenerate tables.** If $u = 0$ or $u = m$ (all failures or all successes
in one group), the test still runs but the confidence interval may be one-sided
or very wide. The same applies when $v = 0$ or $v = n$. These are not errors;
they reflect genuine uncertainty given the data.

**Extreme odds ratios.** Testing a null odds ratio far from the data (e.g.
`odds_ratio = 100` when the observed OR is near 1) is valid but may push the
bisection searches to their limits. Set `precision` smaller (e.g. `1e-4`) if
the returned confidence limits look imprecise, and check that the returned
p-value and confidence interval are still consistent.

**Very small tables.** The algorithm requires that $m + n \geq 2$. For very
small tables (e.g. $m = n = 2$), the test frame may have few rows and the
optimised $\gamma_0$ may be 0 (no borderline rejections), making the modified
test identical to the conservative test. Check `result$gamma0` and
`result$support.data` to confirm the threshold is non-trivial.
 
**Very large tables.** Computation time grows with $(m+1)(n+1)$ because the
rejection matrix has that many entries. The `power = FALSE` and `conf_int =
FALSE` flags skip optional steps and reduce runtime when only the p-value is
needed.

**Grid density and size maximisation.** The dominant cost in most calls is
the maximisation of the local size over $\pi_1$. This is controlled by the
`maze` and `zoom_iter` arguments (for `method = "zoom"`). A denser $\pi_1$ grid
(equivalent to a larger $N$ in the paper's notation) gives a more reliable size
estimate but takes proportionally longer: at $m = 65$, $n = 71$ the function
takes roughly 1.5 seconds at $N = 20$ grid points and around 20 seconds at $N =
300$. The default settings (`maze = 10`, `zoom_iter = 6`) are adequate for most
purposes.
 
**Precision.** The default `precision = 1e-3` is adequate for most purposes.
Results are reported to three decimal places, matching the paper. For simulation
studies or comparisons requiring finer agreement, use `precision = 1e-4` or
smaller, at the cost of additional computation.

## Reference
 
van der Meulen EA, Raymond K, van der Meulen PJ (2021). *Consistent Confidence
Limits, P Values, and Power of the Non-Conservative, Size-α Modified Fisher
Exact Test.* Journal of Biostatistics and Biometric Applications 6(1):102.
